Microlensing as a possible probe of event-horizon structure in quasars

### 2016-10-10 paper1/ms.tex
Prasenjit's new paragraphs

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, refers to sharp changes in the
observed brightness of quasars that have been lensed by an intervening
galaxies, without any changes in the intrinsic luminosity.  It is a
consequence of two things: the very small size of the central engine,
and granularity of the mass distribution of a lensing galaxy due to
stars.  The latter means that the local magnification is not a smooth
function of source position. It contains a complicated network of
singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence, their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  Quasar microlensing is mostly observed in the optical
continuum \citep[e.g.,][]{2012A&A...544A..62S} but is also known in
x-rays \citep[e.g.,][]{2016arXiv160207601N} and has been argued for at
cm wavelengths \citep{2000A&A...358..793K}.  The effect supplies an
upper limit on the size of the central engine
\citep[e.g.,][]{2015ApJ...814L..26M} and can be used to study the mass
distribution and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Microlensing caustics have an
additional remarkable feature: though they can be very complicated,
they have some universal properties well-known from catastrophe
theory.  In particular, very close to the simplest caustics (known as
folds), the magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

In Section~\ref{sec:source-models} we introduce the three source
profiles used in our subsequent models and simulations: a
constant-brightness disc, a circular Gaussian, and the crescent source
introduced by \citep{2013MNRAS.434..765K}.  The latter is simply a
constant-brightness disc with a smaller, non-concentric disc cut out
of it.  We also derive the half-light radius for a crescent.  The
half-light radius can characterise the source size for all three types
of source.

@@

In the case of a high-quality lightcurve with 
insignificant noise and measurement errors, 
the parameters can be obtained 
by simply identifying the characteristic 
instances of time without making use of 
the actual values of the magnification
 map. If the effect of the errors and the 
noise distorts the magnification time function enough so that the 
characteristic epochs are not identifiable with the characteristic 
periods still visible, the boundaries of the periods 
can be roughly estimated. Furthermore, if direct estimates of 
the parameters cannot be obtained we propose the use of a strong
statistical tool such as Markov-Chain Monte Carlo.
 In Section 6  we have studied the possibility of identifying a crescent 
source and the possibility of recovering the 
respective parameters. As magnification map, we have used a complex numerical 
one generated with the microlensing code by \cite{1999A&A...346L...5W}. 
In addition, we have added to the signal a Gaussian 
noise with an SNR of 1.6.  In the experiment,
 we have considered all nine combinations of
 original source profiles and assumed fitting source models.
Effectively we have fitted using MCMC all three sources with all three assumed fitting models. The results of the experiment allowed us to build a 
procedure for distinguishing the shape of the source assuming that one of the three models we have considered is a good approximation.  
The procedure would be useful for observers that endeavour to gain more information about an unresolved source for which they can study 
the microlensing lightcurve.  As a first step in the procedure, one should first attempt to use MCMC with a Gaussian model assumption. 
If either the value of the $\chi^2$ or the number of rejected datapoints is large then the next step is to change the assumed source model to 
a uniform disc and redo the fitting. Finally, if the uniform disc assumption is rejected as well then the fitting should be done with a 
crescent source model assumption. If at each of the three steps the data rejects the model then the source cannot be approximated by any of the three models. 
Otherwise, if at one of the steps the data does not reject the model then the respective model is a good approximation.

>>>>

%Gravitational microlensing of quasars, reviewed in
%Section~\ref{sec:microlensing} below, refers to sharp changes in the
%observed brightness of quasars that have been lensed by an intervening
%galaxies, without any changes in the intrinsic luminosity.

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, is a phenomenon seen in quasars
that have been multiply imaged by galaxies.  Individual images in a
multiple-image system can undergo sharp and seemingly random
brightness changes.  This can occur, even for constant intrinic
luminosity, as a consequence of two things: It is a
consequence of two things: the very small size of the central engine,
and granularity of the mass distribution of a lensing galaxy due to
stars.  The latter means that the local magnification is not a smooth
function of source position. It contains a complicated network of
singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence, their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  Quasar microlensing is mostly observed in the optical
continuum \citep[e.g.,][]{2012A&A...544A..62S} but is also known in
x-rays \citep[e.g.,][]{2016arXiv160207601N} and has been argued for at
cm wavelengths \citep{2000A&A...358..793K}.  The effect supplies an
upper limit on the size of the central engine
\citep[e.g.,][]{2015ApJ...814L..26M} and can be used to study the mass
distribution and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Microlensing caustics have an
additional remarkable feature: though they can be very complicated,
they have some universal properties well-known from catastrophe
theory.  In particular, very close to the simplest caustics (known as
folds), the magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

%In Section~\ref{sec:source-models} we introduce the three source
%profiles used in our subsequent models and simulations: a
%constant-brightness disc, a circular Gaussian, and the crescent source
%introduced by \citep{2013MNRAS.434..765K}.  The latter is simply a
%constant-brightness disc with a smaller, non-concentric disc cut out
%of it.  We also derive the half-light radius for a crescent.  The
%half-light radius can characterise the source size for all three types
%of source.

In Section~\ref{sec:numerics} we carry out source fitting to
lightcurves, with both noise and systematic errors are present.  We
generate a mock lightcurves by taking a crescent source across the
path AB in Figure~\ref{fig:magnification_map}, and then adding noise.
The path simulates crossing a clean but not ideal fold.  We then fit
this lightcurve to templates from both crescent and circular-Gaussian
sources across the path CB.  That is, the templates come from a
similar but not identical caustic, thus deliberately generating a
systematic error.  Despite this systematic error, a crescent source
fits the data, while a Gaussian source is rejected from the $\chi^2$
value.

@@

%In the case of a high-quality lightcurve with 
%insignificant noise and measurement errors, 
%the parameters can be obtained 
%by simply identifying the characteristic 
%instances of time without making use of 
%the actual values of the magnification
% map. If the effect of the errors and the 
%noise distorts the magnification time function enough so that the 
%characteristic epochs are not identifiable with the characteristic 
%periods still visible, the boundaries of the periods 
%can be roughly estimated. Furthermore, if direct estimates of 
%the parameters cannot be obtained we propose the use of a strong
%statistical tool such as Markov-Chain Monte Carlo.
% In Section 6  we have studied the possibility of identifying a crescent 
%source and the possibility of recovering the 
%respective parameters. As magnification map, we have used a complex numerical 
%one generated with the microlensing code by \cite{1999A&A...346L...5W}. 
%In addition, we have added to the signal a Gaussian 
%noise with an SNR of 1.6.  In the experiment,
% we have considered all nine combinations of
% original source profiles and assumed fitting source models.
%Effectively we have fitted using MCMC all three sources with all three assumed fitting models. The results of the experiment allowed us to build a 
%procedure for distinguishing the shape of the source assuming that one of the three models we have considered is a good approximation.  
%The procedure would be useful for observers that endeavour to gain more information about an unresolved source for which they can study 
%the microlensing lightcurve.  As a first step in the procedure, one should first attempt to use MCMC with a Gaussian model assumption. 
%If either the value of the $\chi^2$ or the number of rejected datapoints is large then the next step is to change the assumed source model to 
%a uniform disc and redo the fitting. Finally, if the uniform disc assumption is rejected as well then the fitting should be done with a 
%crescent source model assumption. If at each of the three steps the data rejects the model then the source cannot be approximated by any of the three models. 
%Otherwise, if at one of the steps the data does not reject the model then the respective model is a good approximation.

In the case of a high-quality lightcurve, the parameters of a crescent
source can be estimated by simply identifying signatures on the
lightcurve corresponding to transits of the particular source features
across the caustic.  This simple idea can be supplemented with
statistics.  We do so in Section~\ref{sec:numerics}, where we take a
realistic magnification map generated by the microlensing code
from \cite{1999A&A...346L...5W}, compute a caustic-crossing lightcurve
from it, and then fit the lightcurve with other caustic-crossing
templates.  The conclusion is that a crescent source can be
distinguised from a circular source, on the basis of caustic-crossing
lightcurves, if the caustic is a clean but not ideal fold.
